### 2019-01-15 submission/manuscript.tex
Add inline R code and refs to figures

\textbf{\emph{Sequencing errors vary by the number of cycles and the DNA
polymerase used in PCR.}} The presence of sequence errors can confound
the ability to accurately classify 16S rRNA gene sequences and group
sequences into operational taxonomic units (OTUs). More importantly,
sequencing errors themselves can alter the representation of the
community. Therefore, it is important to minimize the number of
sequencing errors. Using a widely-used approach that generates the
lowest reported error rate, we quantified the error rate by sequencing
the V4 region of the 16S rRNA genes from an 8 member mock community. We
also removed any contigs that were at least three bases more similar to
a chimera of two references than to a single reference sequence
{[}ref{]}. Regardless of the polymerase, the error rate increased with
the number of rounds of amplification. Using 30 rounds of PCR is a
common approach across diverse types of samples. Among the data
generated using 30 rounds of PCR the Accuprime polymerase had the
highest error rate (i.e.~0.XXX\%) followed by the Platinum
(i.e.~0.XXX\%), Phusion (i.e.~0.XXX\%), Kappa (i.e.~0.XXX\%), and Q5
(i.e.~0.XXX\%) polymerases (Figure 1). When we applied the
pre.clustering denoising step, which merges the counts of reads within 2
nt of a more abundant sequence, the error rates dropped considerably
such that the Platinum polymerase had the highest error rate
(i.e.~0.XXX\%) followed by the Accuprime (i.e.~0.XXX\%), Phusion
(i.e.~0.XXX\%), Kappa (i.e.~0.XXX\%), and Q5 (i.e.~0.XXX\%) polymerases
(Figure 1). Although specific recommendations are difficult to make
because the impact of the actual community structure and concentration
of the initial DNA template are likely to have an impact on the results,
it is clear that using as few PCR cycles as necessary and a polymerase
with the lowest possible error rate is a good guide to minimizing the
impact of polymerase on the error rate.

\textbf{\emph{The fraction of sequences identified as being chimeric
varies by the number of cycles and the DNA polymerase used in PCR.}}
Chimeric PCR products can significantly confound downstream analyses.
Although numerous bioinformatic tools exist to identify and remove
chimeric sequences with high specificity, their sensitivity is
relatively low and can be reduced by the presence of sequencing errors.
We identified those contigs that were at least three bases more similar
to a chimera of two references than to a single reference sequence. As
expected from previous studies, the number of chimeras increased with
rounds of amplification. Interestingly, the fraction of chimeric
sequences from the mock community varied by the type of polymerase used.
After 30 rounds of PCR, the Platinum polymerase had the highest chimera
rate (i.e.~XX\%) followed by the Q5 (i.e.~XX\%), Phusion (i.e.~XX\%),
Kappa (i.e.~XX\%), and Accuprime (i.e.~XX\%) polymerases. We used the
UChime algorithm to detect chimeras in our pre-clustered mock community
data and calculate the algorithm's sensitivity and specificity. For all
polymerases, the specificity was above XX\% and showed a weak
association with the number of cycles used. There was considerable
inter-polymerase and inter-round of amplification variation in the
sensitivity of UChime to detect the chimeras from the mock community.
This suggested that the residual error rate after pre-clustering the
sequence data did not compromise the sensitivity. In general, the
sensitivity of UChime varied between XX and XX\%. The generalizability
of these results is limited because we used a single mock community with
limited genetic diversity. Although we did not know the true chimera
rate for our four human stool samples, we were able to calculate the
fraction of sequences that UChime identified as being chimeric. These
results followed those from the mock communities; additional rounds of
amplification significantly increased the rate of chimeras and there was
variation between the polymerases that we used. Although it was not
possible to identify the features of a polymerase that resulted in
higher rates of chimeras, it is clear that using the smallest number of
PCR cycles possible will minimize the impacts of chimeras.

\textbf{\emph{At the sequence level, PCR amplification bias is subtle.}}
Since researchers began using PCR to sequence 16S rRNA genes there has
been concern that amplifying fragments from a mixed template pool could
lead to a biased representation in the pool of products and would
confound downstream analyses. The mock community was generated by mixing
equal amounts of genomic DNA from each of the 8 bacteria resulting in
uneven representation of the \emph{rrn} operons across the bacteria. The
vendor of the mock community subjects each lot of genomic DNA to shotgun
sequencing to more accurately quantify the actual abundance of each
organism in the community. We compared the expected relative abundance
of the 16S rRNA genes from each bacterium in the mock community to the
data we generated across rounds of amplification and polymerase.
Interestingly, for some bacteria, their representation became less
biased with additional rounds of PCR (e.g.~\emph{L. fermentum}), while
others became more biased (e.g.~\emph{E. faecalis}), and others had
little change (e.g.~\emph{B. subtilis}). The percentage of bases in the
V4 region that were guanines or cytosines was not predictive of the
amount of bias. Across the strains there was no variation in the length
of their V4 regions and they each had the same sequence in the region
that the primers annealed. One of the bacteria represented in the mock
community, \emph{S. enterica}, had 6 identical copies of the V4 region
and 1 copy that differed from those by one nucleotide. The dominant copy
had a thymidine and the rare copy had a guanine. We used the sequence
data to calculate the ratio of the dominant to rare variants from
\emph{S. enterica} (Figure. XXX). The Accuprime, Phusion, Platinum, and
Q5 polymerases converged to a ratio of 5.XXX; however, the ratio for the
Kappa polymerase varied between 6.XXX and 7.XXX for 25 to 35 rounds of
PCR. Given the subtle nature of the variation in the relative abundances
of each 16S rRNA gene fragment, it was not possible to create
generalizable rules that would explain the bias and the previous
observation regarding the guanine and cytosine relative abundance was
not observed with our data.

\textbf{\emph{At the community level, the effects of PCR amplification
bias grow with additional rounds of PCR.}} Because the distance between
samples and across rounds of PCR could be artificially inflated due to
sequencing errors and chimeras, we used analyzed the alpha and beta
diversity of the mock community data at different phases of the sequence
curation pipeline. First, we removed the chimeras from the mock
community data as described above and mapped the individual reads to the
OTUs that the 16S rRNA gene fragments would cluster into if there were
no sequencing errors. This gave us a community distribution that
reflected the distribution following PCR without any artifacts. Alpha
diversity metrics such as richness and Shannon and Inverse Simpson
diversity metrics increased with the number of rounds of PCR for all
polymerases except for the Kappa polymerase, where the richness remained
constant but the diversity decreased. These data suggest that PCR had
the effect of making the community distribution more even than it was
originally. Next, we used the true sequence diversity, but removed
chimeras based on mapping reads to all possible chimeras between
reference sequences, and clustered the reads to OTUs. The same trends
were observed as with perfect sequence data except that the richness and
diversity metrics trended higher. Finally, we used the observed sequence
data and used the UCHIME algorithm to identify chimeras. Again, we
observed the same trends as with the other analyses except the richness
and diversity metrics trended higher. These comparisons demonstrated
that although the bias at the sequence level was subtle, PCR does have a
bias that is exacerbated by errors and chimeras. When we measured the
Bray-Curtis distance between the communities observed after 25 rounds of
amplification and those at 30 and 35 rounds of amplification the
distances between 25 and 35 rounds were higher than between 25 and 30
rounds for each of the polymerases by an average of 0.XX units. The
Platinum polymerase varied the most across rounds of amplification (25
vs 30 rounds: 0.XX; 25 vs 35 rounds: 0.XX). Although the distances
between samples were small, the ordination of these distances showed a
clear change in community structure with increasing rounds of PCR. This
observation was supported by our statistical analysis, which revealed
that the number of rounds of PCR (R\textsuperscript{2}=0.XX, P=0.XXXX)
was a larger factor than the choice of polymerase
(R\textsuperscript{2}=0.XX, P=0.XXXX). These results demonstrate that
subtle differences in relative abundances can have an impact on overall
community structure. This variation underscores the importance of only
comparing sequence data that have been generated using the same PCR
conditions.

\textbf{\emph{The choice of polymerase or the number of rounds of
amplification have little impact on the relative interpretation of
community-wide metrics of diversity.}} The biases that we observed at
the population and community levels using mock community data appeared
to be small relative to the expected differences between biological
samples. To study this further, we calculated alpha and beta-diversity
metrics using the human stool samples for each of the polymerases and
rounds of amplification. We calculated the number of observed OTUs,
Shannon diversity, inverse Simpson diversity index for each condition
and donor. Although there were clear differences between conditions, the
relative ordering of the stool samples did not meaningfully vary across
conditions. When we characterized the variation between rounds of
amplification using human stool samples, the distance between the 25 and
30 rounds and 25 and 35 rounds varied considerably between samples and
polymerases. In general the inter-round variation was lowest for the
data generated using the Kappa and Accuprime polymerases. The data
generated using the Platinum polymerase was consistent across rounds,
but it was more biased than these polymerases. Considering the average
distance across the four samples varied between 0.XX and 0.XX,
regardless of the polymerases and number of rounds of amplification, any
bias due to amplification is unlikely to obscure community-wide
differences between samples. In support of this was our principle
coordinates analysis of the Bray-Curtis distances, which revealed
distinct clusters by donor. Within each cluster there were no obvious
patterns related to the polymerase or number of rounds of PCR. However,
our statistical analysis using adonis revealed statistically significant
differences in the community structures with the subject explaining the
most variation (R\textsuperscript{2}=0.XXX, P=0.XXXX), followed by the
number of rounds of PCR (R\textsuperscript{2}=0.XXX, P=0.XXXX) and the
choice of polymerase (R\textsuperscript{2}=0.XXX, P=0.XXXX). Together,
these results indicate that for a coarse analysis of communities, the
choice of number of rounds of amplification or polymerase are not
important, but that they must be consistent across samples. It is
difficult to develop a specific recommendation based on the level of
bias across rounds of PCR or polymerases; however, the general
suggestion is to use as few rounds of amplification as possible.

\textbf{\emph{There is little evidence of a relationship between
polymerase or number of rounds of amplification on PCR drift.}} There
have been concerns that the same template DNA subjected to the same PCR
conditions could result in different representations of communities
because of random drift. To test this, we determined the average
Bray-Curtis distance between replicate reactions using the same
polymerase and number of rounds of amplification. Using the mock
community data there were no obvious trends. The average Bray-Curtis
distance within a set of conditions varied by 0.XX to 0.XX units.
Although we did not obtain replicates of each of the stool samples, the
intra-sample variation for each set of conditions was consistent and
varied between 0.XX and 0.XX units. These data suggest that amplicon
sequencing is robust to random variation in amplification and that any
differences are likely to be smaller than what is considered
biologically relevant.

@@

\textbf{\emph{PCR protocol.}} Five high fidelity DNA polymerases were
tested including AccuPrime\textsuperscript{TM} (ThermoFisher, MA, USA),
KAPA HIFI (Roche, IN, USA), Phusion (New England Biolabs, MA, USA),
Platinum (ThermoFisher, MA, USA), and Q5 (New England Biolabs, MA, USA).
Manufacturer recommendations were followed except for the annealing and
extension times, which were selected based on previously published
protocols (18, 37). Primers targeting the V4 region of the 16S rRNA gene
were used with modifications to generate MiSeq amplicon libraries
{[}@Kozich2013;
\url{https://github.com/SchlossLab/MiSeq_WetLab_SOP/}{]}. The number of
rounds of PCR used for each sample and polymerase 15 and increased by 5
rounds up to 35 cycles. Insufficient PCR product was generated using 15
rounds and has not been included in our analysis.

\textbf{\emph{Library generation and sequencing.}} Each PCR condition
(i.e.~combination of polymerase and number of rounds of PCR) were
replicated four times for the mock community and one time for each fecal
sample. Libraries were generated as described previously {[}@Kozich2013;
\url{https://github.com/SchlossLab/MiSeq_WetLab_SOP/}{]}. The libraries
were sequenced using the Illumina MiSeq sequencing platform to generate
paired 250-nt reads.

>>>>

\textbf{\emph{Sequencing errors vary by the number of cycles and the DNA
polymerase used in PCR.}} The presence of sequence errors can confound
the ability to accurately classify 16S rRNA gene sequences and group
sequences into operational taxonomic units (OTUs). More importantly,
sequencing errors themselves can alter the representation of the
community. Therefore, it is important to minimize the number of
sequencing errors. Using a widely-used approach that generates the
lowest reported error rate, we quantified the error rate by sequencing
the V4 region of the 16S rRNA genes from an 8 member mock community. We
also removed any contigs that were at least three bases more similar to
a chimera of two references than to a single reference sequence
{[}ref{]}. Regardless of the polymerase, the error rate increased with
the number of rounds of amplification. Using 30 rounds of PCR is a
common approach across diverse types of samples. Among the data
generated using 30 rounds of PCR the Accuprime polymerase had the
highest error rate (i.e.~0.124\%) followed by the Platinum
(i.e.~0.094\%), Phusion (i.e.~0.064\%), Kappa (i.e.~0.062\%), and Q5
(i.e.~0.060\%) polymerases (Figure 1). When we applied the
pre.clustering denoising step, which merges the counts of reads within 2
nt of a more abundant sequence, the error rates dropped considerably
such that the Platinum polymerase had the highest error rate
(i.e.~0.014\%) followed by the Accuprime (i.e.~0.012\%), Q5
(i.e.~0.0053\%), Phusion (i.e.~0.0049\%), and Kappa (i.e.~0.0049\%)
polymerases (Figure 1). Although specific recommendations are difficult
to make because the impact of the actual community structure and
concentration of the initial DNA template are likely to have an impact
on the results, it is clear that using as few PCR cycles as necessary
and a polymerase with the lowest possible error rate is a good guide to
minimizing the impact of polymerase on the error rate.

\textbf{\emph{The fraction of sequences identified as being chimeric
varies by the number of cycles and the DNA polymerase used in PCR.}}
Chimeric PCR products can significantly confound downstream analyses.
Although numerous bioinformatic tools exist to identify and remove
chimeric sequences with high specificity, their sensitivity is
relatively low and can be reduced by the presence of sequencing errors.
We identified those contigs that were at least three bases more similar
to a chimera of two references than to a single reference sequence. As
expected from previous studies, the number of chimeras increased with
rounds of amplification. Interestingly, the fraction of chimeric
sequences from the mock community varied by the type of polymerase used.
After 30 rounds of PCR, the Platinum polymerase had the highest chimera
rate (i.e.~18.2\%) followed by the Q5 (i.e.~8.1\%), Phusion
(i.e.~7.5\%), Kappa (i.e.~2.3\%), and Accuprime (i.e.~0.9\%)
polymerases. We used the UChime algorithm to detect chimeras in our
pre-clustered mock community data and calculate the algorithm's
sensitivity and specificity. For all polymerases, the specificity was
above 95.8\% and showed a weak association with the number of cycles
used. There was considerable inter-polymerase and inter-round of
amplification variation in the sensitivity of UChime to detect the
chimeras from the mock community. This suggested that the residual error
rate after pre-clustering the sequence data did not compromise the
sensitivity. The sensitivity of UChime varied between 50 and 84.8\%. The
generalizability of these results is limited because we used a single
mock community with limited genetic diversity. Although we did not know
the true chimera rate for our four human stool samples, we were able to
calculate the fraction of sequences that UChime identified as being
chimeric. These results followed those from the mock communities;
additional rounds of amplification significantly increased the rate of
chimeras and there was variation between the polymerases that we used.
Although it was not possible to identify the features of a polymerase
that resulted in higher rates of chimeras, it is clear that using the
smallest number of PCR cycles possible will minimize the impacts of
chimeras.

\textbf{\emph{At the sequence level, PCR amplification bias is subtle.}}
Since researchers began using PCR to sequence 16S rRNA genes there has
been concern that amplifying fragments from a mixed template pool could
lead to a biased representation in the pool of products and would
confound downstream analyses. The mock community was generated by mixing
equal amounts of genomic DNA from each of the 8 bacteria resulting in
uneven representation of the \emph{rrn} operons across the bacteria. The
vendor of the mock community subjects each lot of genomic DNA to shotgun
sequencing to more accurately quantify the actual abundance of each
organism in the community. We compared the expected relative abundance
of the 16S rRNA genes from each bacterium in the mock community to the
data we generated across rounds of amplification and polymerase.
Interestingly, for some bacteria, their representation became less
biased with additional rounds of PCR (e.g.~\emph{L. fermentum}), while
others became more biased (e.g.~\emph{E. faecalis}), and others had
little change (e.g.~\emph{B. subtilis}). The percentage of bases in the
V4 region that were guanines or cytosines was not predictive of the
amount of bias. Across the strains there was no variation in the length
of their V4 regions and they each had the same sequence in the region
that the primers annealed. One of the bacteria represented in the mock
community, \emph{S. enterica}, had 6 identical copies of the V4 region
and 1 copy that differed from those by one nucleotide. The dominant copy
had a thymidine and the rare copy had a guanine. We used the sequence
data to calculate the ratio of the dominant to rare variants from
\emph{S. enterica} (Figure XXX). The Accuprime, Phusion, Platinum, and
Q5 polymerases converged to a ratio of 5.4; however, the ratio for the
Kappa polymerase varied between 6.1 and 6.8 for 25 to 35 rounds of PCR.
Given the subtle nature of the variation in the relative abundances of
each 16S rRNA gene fragment, it was not possible to create generalizable
rules that would explain the bias and the previous observation regarding
the guanine and cytosine relative abundance was not observed with our
data.

\textbf{\emph{At the community level, the effects of PCR amplification
bias grow with additional rounds of PCR.}} Because the distance between
samples and across rounds of PCR could be artificially inflated due to
sequencing errors and chimeras, we used analyzed the alpha and beta
diversity of the mock community data at different phases of the sequence
curation pipeline. First, we removed the chimeras from the mock
community data as described above and mapped the individual reads to the
OTUs that the 16S rRNA gene fragments would cluster into if there were
no sequencing errors. This gave us a community distribution that
reflected the distribution following PCR without any artifacts. Alpha
diversity metrics such as richness and Shannon and Inverse Simpson
diversity metrics increased with the number of rounds of PCR for all
polymerases except for the Kappa polymerase, where the richness remained
constant but the diversity decreased. These data suggest that PCR had
the effect of making the community distribution more even than it was
originally. Next, we used the true sequence diversity, but removed
chimeras based on mapping reads to all possible chimeras between
reference sequences, and clustered the reads to OTUs. The same trends
were observed as with perfect sequence data except that the richness and
diversity metrics trended higher. Finally, we used the observed sequence
data and used the UCHIME algorithm to identify chimeras. Again, we
observed the same trends as with the other analyses except the richness
and diversity metrics trended higher. These comparisons demonstrated
that although the bias at the sequence level was subtle, PCR does have a
bias that is exacerbated by errors and chimeras. When we measured the
Bray-Curtis distance between the communities observed after 25 rounds of
amplification and those at 30 and 35 rounds of amplification the
distances between 25 and 35 rounds were higher than between 25 and 30
rounds for each of the polymerases by an average of 0.023 units. The
Platinum polymerase varied the most across rounds of amplification (25
vs 30 rounds: 0.13; 25 vs 35 rounds: 0.16). Although the distances
between samples were small, the ordination of these distances showed a
clear change in community structure with increasing rounds of PCR. This
observation was supported by our statistical analysis, which revealed
that the number of rounds of PCR (R\textsuperscript{2}=0.20,
P\textless{}0.001) was a larger factor than the choice of polymerase
(R\textsuperscript{2}=0.20, P\textless{}0.001). These results
demonstrate that subtle differences in relative abundances can have an
impact on overall community structure. This variation underscores the
importance of only comparing sequence data that have been generated
using the same PCR conditions.

\textbf{\emph{The choice of polymerase or the number of rounds of
amplification have little impact on the relative interpretation of
community-wide metrics of diversity.}} The biases that we observed at
the population and community levels using mock community data appeared
to be small relative to the expected differences between biological
samples. To study this further, we calculated alpha and beta-diversity
metrics using the human stool samples for each of the polymerases and
rounds of amplification. We calculated the number of observed OTUs,
Shannon diversity, inverse Simpson diversity index for each condition
and donor. Although there were clear differences between conditions, the
relative ordering of the stool samples did not meaningfully vary across
conditions. When we characterized the variation between rounds of
amplification using human stool samples, the distance between the 25 and
30 rounds and 25 and 35 rounds varied considerably between samples and
polymerases. In general the inter-round variation was lowest for the
data generated using the Kappa and Accuprime polymerases. The data
generated using the Platinum polymerase was consistent across rounds,
but it was more biased than these polymerases. Considering the average
distance across the four samples varied between 0.39 and 0.56,
regardless of the polymerases and number of rounds of amplification, any
bias due to amplification is unlikely to obscure community-wide
differences between samples. In support of this was our principle
coordinates analysis of the Bray-Curtis distances, which revealed
distinct clusters by donor. Within each cluster there were no obvious
patterns related to the polymerase or number of rounds of PCR. However,
our statistical analysis using adonis revealed statistically significant
differences in the community structures with the subject explaining the
most variation (R\textsuperscript{2}=0.79, P\textless{}0.001), followed
by the number of rounds of PCR (R\textsuperscript{2}=0.044,
P\textless{}0.001) and the choice of polymerase
(R\textsuperscript{2}=0.033, P\textless{}0.001). Together, these results
indicate that for a coarse analysis of communities, the choice of number
of rounds of amplification or polymerase are not important, but that
they must be consistent across samples. It is difficult to develop a
specific recommendation based on the level of bias across rounds of PCR
or polymerases; however, the general suggestion is to use as few rounds
of amplification as possible.

\textbf{\emph{There is little evidence of a relationship between
polymerase or number of rounds of amplification on PCR drift.}} There
have been concerns that the same template DNA subjected to the same PCR
conditions could result in different representations of communities
because of random drift. To test this, we determined the average
Bray-Curtis distance between replicate reactions using the same
polymerase and number of rounds of amplification. Using the mock
community data there were no obvious trends. The average Bray-Curtis
distance within a set of conditions varied by 0.063 to 0.11 units.
Although we did not obtain replicates of each of the stool samples, the
intra-sample variation for each set of conditions was consistent and
varied between 0.50 and 0.56 units. These data suggest that amplicon
sequencing is robust to random variation in amplification and that any
differences are likely to be smaller than what is considered
biologically relevant.

@@

\textbf{\emph{PCR protocol.}} Five high fidelity DNA polymerases were
tested including AccuPrime\textsuperscript{TM} (ThermoFisher, MA, USA),
KAPA HIFI (Roche, IN, USA), Phusion (New England Biolabs, MA, USA),
Platinum (ThermoFisher, MA, USA), and Q5 (New England Biolabs, MA, USA).
Manufacturer recommendations were followed except for the annealing and
extension times, which were selected based on previously published
protocols (18, 37). Primers targeting the V4 region of the 16S rRNA gene
were used with modifications to generate MiSeq amplicon libraries (18)
(\url{https://github.com/SchlossLab/MiSeq_WetLab_SOP/}). The number of
rounds of PCR used for each sample and polymerase 15 and increased by 5
rounds up to 35 cycles. Insufficient PCR product was generated using 15
rounds and has not been included in our analysis.

\textbf{\emph{Library generation and sequencing.}} Each PCR condition
(i.e.~combination of polymerase and number of rounds of PCR) were
replicated four times for the mock community and one time for each fecal
sample. Libraries were generated as described previously (18)
(\url{https://github.com/SchlossLab/MiSeq_WetLab_SOP/}). The libraries
were sequenced using the Illumina MiSeq sequencing platform to generate
paired 250-nt reads.
